scTOP: physics-inspired order parameters for cellular identification and visualization of developmental trajectories

### 2022-11-30 scTOPv2.tex
manuscript version 2

\author{Maria Yampolskaya}
\email{mariay@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\author{Michael Herriges}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\author{Laertis Ikonomou}
\affiliation{Department of Oral Biology. University at Buffalo, The State University of New York, Buffalo, NY, USA}
\affiliation{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
\author{Darell Kotton}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}

\author{Pankaj Mehta}
\email{pankajm@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
\affiliation{Biological Design Center, Boston University, Boston, MA 02215, USA}

@@

\section{Introduction}
Humans and other mammals have hundreds of specialized cells that arise through a complex developmental process  [CITE]. At early embryonic stages, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers[CITE]. However, as embryos mature, cellular identity becomes less plastic and more specified through a complex spatiotemporal process involving signaling and environmental cues. The questions of how and why cells end up in one fate or another is a fundamental question in developmental biology [CITE]. More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineer cells \emph{in vitro}, with important clinical application for human health and disease [CITE].

@@

An ideal analysis algorithm for scRNA-seq data should possess certain desirable properties. 
First, the algorithm should incorporate explicit assumptions that can be kept consistent between analyses. Second, the algorithm should be deterministic, further enhancing reproducibility. Third, the axes in visualization should be interpretable, intuitive, and remain fixed across analysis and datasets. Fourth, the algorithm should measure cell identity in a continuous way instead of assigning discrete identity labels, thus enabling a more subtle and fine-grained analysis of cellular identity and developmental trajectories. Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single cell atlases now being generated [].

@@

\subsubsection{Example 1: lung lineages}
To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al. \cite{herriges_durable_2022} onto the mouse cell atlas. The Herriges et al. dataset  measures gene expression levels of six different specialized lung types from mice used as control samples. See figure \ref{FIG:2} (a). The reference basis used to find the individual scTOP scores for this data set was derived from the Mouse Cell Atlas. Despite the fact that the Kotton lab dataset and the Mouse Cell Atlas were created in different laboratories and different sequencing methods, the resulting scTOP scores distinctly separate the cells from the Kotton laboratory. The resulting clusters agree with the annotations of the true cell type provided by the Kotton laboratory.

@@

The mouse brain atlas\cite{yao_taxonomy_2021} sequenced over one million cells and clustered them into 42 subclasses and 101 supertypes. For a reference basis, we reserved a subset of 200 cells from each of the identified subclasses to use as training data. The test data, from which the accuracy score was calculated, was the remaining cells of each type not included in the training data. Ten different cell types are visualized in a grid in figure \ref{FIG:2} (b). The color of each grid square corresponds to the average scTOP score of cells annotated as the type on the y-axis, with the cell type corresponding to the scTOP dimension on the x-axis. As shown by the distinct blue diagonal, the average scTOP score correctly matches the corresponding annotated cell type.

@@

scTOP correctly identifies individual cells, and it can also identify the tissue to which a population of cells belongs. As shown in \ref{FIG:2} (c), we found the aggregate scTOP scores for cells from the Mouse Cell Atlas projected onto the reference basis of Tabula Muris organs \cite{schaum_single-cell_2018}. Most cell types were correctly and strongly identified with the annotated tissue of origin. Parenchymal cell types were the most associated with the correct organ, while stromal cells were sometimes misidentified -- for example, the mammary gland endothelial cells. This is consistent with the fact that stromal cells are less specified to each organ. A discussion of the accuracy of scTOP when applied to stromal types versus parenchymal types may be found in appendix \ref{failure_cases}.

\begin{figure}
	\centering
		\includegraphics[scale=0.5]{figs/scTOP manuscript fig 2.pdf}
	\caption{scTOP identifies mouse cell types. (a) Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on scTOP score axes. The data points correspond to individual cells, and in the first row of plots, the marker shape and color indicate the "true type" as determined by annotations from Herriges et al. The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types I and II. The second, third, and fourth rows of plots have individual cells colored by the marker genes for the type indicated by the color bars at the end of each row. (b) A subset of cells from the Mouse Brain Atlas are used as the reference basis to query other cells from the same data set. The shade of each bin corresponds to the average scTOP score for individual cells of the type indicated on the y-axis, compared to the reference types on the x-axis. The diagonal indicates that scTOP accurately matches query cells to the true reference type. (c) Aggregate samples from the Mouse Cell Atlas are compared to a reference basis consisting of organs from Tabula Muris. The scTOP scores are significantly higher when comparing a cell type with the organ of origin. This coarse-grain identification shows scTOP is able to classify tissues at different resolutions.}
	\label{FIG:2}
\end{figure}

@@

The CellBench dataset \cite{tian_benchmarking_2019} was designed specifically to benchmark scRNA-seq analysis methods. We used the subset of data pertaining to five cancer lines. 200 cells from each line were used as training data to create the reference basis, then the rest of the cells were input as query data. As shown in figure \ref{FIG:3} (a) and table \ref{table:1}, scTOP classifies cell line identity of the test data with 100 percent accuracy. {\color{red} PM: How did other methods do?}

To further prove the power of scTOP in classifying human samples, we analyzed data from the human lung atlas dataset \cite{travaglini_molecular_2020}. This dataset sequenced thousands of human lung cells and identified 58 phenotypic popoulations. For our analysis, we restricted the training and test data to epithelial and stromal types that had at least 200 cells in each type population. The reference basis was created using 80\% of the cells from each population, and the query data consisted of the remaining 20\% of the data. Figure \ref{FIG:3} (b) shows that the scTOP scores were consistently high in cases where the score type matched the true type of the query.

\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/fig3.pdf}
	\caption{scTOP correctly identifies human tissues. (a) The CellBench data set contains human cell cancer lines specifically processed to serve as benchmarking for scRNA-seq algorithms. We take a subset of cells to use as the reference basis in analyzing the rest of the CellBench data. The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis. The y-axis lists the true types of the queried cells. The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types. (b) Similar to (a), cells from the Human Lung Atlas are compared to a reference basis constructed from cells from the Human Lung Atlas. The colors of the bins correspond to average scTOP scores of individual cells. Again, the diagonal is the most prominent section, showing that cells of the y-axis types are correctly matched to the reference x-axis identities.}
	\label{FIG:3}
\end{figure}

@@

\subsubsection{Lung development}
{\color{red} PM: We need to add paragraph with some biological conclusions here -- even if already known sighting relevant literature.. when we see  divergences in lung development ... etc. --- will need final analysis}

The healthy formation of lungs is vital to mammalian life. For a mammal to breathe and live independently after birth, the final stages of embryonic development must equip it with the ability to exchange gases with its environment. Zepp et al. [CITE] sought to examine the formation of pulmonary alveoli, small sacs where oxygen and carbon dioxide are transferred between the lungs and the air, in the developing mouse. They sequenced mouse lungs between post-conception day 12.5 (E12.5) and postnatal day 42 (P42). We found the scTOP scores for cells annotated by Zepp et al. as type I pneumocytes, type II pneumocytes, and epithelial progenitor cells. The reference basis consisted of adult cell types from the Mouse Cell Atlas and an E12 epithelial progenitor cell type from the Atlas of Lung Development \cite{negretti_single-cell_2021}. In \ref{FIG:4}, we show the scTOP scores for the two adult alveolar types using the Mouse Cell Atlas as the reference basis.

{\color{red} MY: Discuss the clear transition to adult cells, then speculate about the co-expressing middle cells (which exist even at P7). Figure 1I from Zepp et al. conjectures that one begins to see separate alveolar type 1 and type 2 populations between E15.5 and E17.5. According to our plot, there are populations of cells clearly headed towards the two types, but there is also a population of cells in between expressing both type 1 and type 2 signatures. Show that it doesn't express Krt8 or the other "transitional cell" genes; it looks more like a spurious attractor/combination of AT1 + AT2}

Because scTOP provides us with reference axes that don't change according to whatever data is included, it is possible to track different data points across time on the same axes. These axes are also easily interpretable: they measure the similarity to a canonical known cell type.

\begin{figure}
	\centering
		\includegraphics[scale=0.75]{figs/scTOP manuscript fig4.pdf}
	\caption{Beyond bench-marking: tracing development in cell-type space with scTOP, using developmental mouse lung data. (a) scTOP illustrates specification of alveolar type I and type II cells in murine embryonic development. Alveolar cells from Zepp et al. are compared to a reference basis constructed from adult lung cells from Herriges et al. and early epithelial cells from the Atlas of Lung Development. The labels next to each data point indicate the age of the mouse embryo from which lungs were extracted, from post-conception day 12 (E12) to day of birth (P0). }
	\label{FIG:4}
\end{figure}

@@

To apply LARRY to blood cells, they extracted hematopoietic progenitor cells from mice, barcoded the cells, waited for the progenitors to multiply in primary culture, then performed scRNA-seq at days 2, 4, and 6 post-barcoding. Then, they extracted blood cells at four and six days after transplantation. The resulting dataset contains over 300,000 cells, with 2,632 clones spanning time points. Clones belonging to the same clonal family are referred to as "sisters." They then used pseudotime orderings to identify a transcriptomic map to identify lienages of the cells. This map used transcriptomic similarity to place the individual cells accoridng to "pseudotime". With scTOP, creating a pseudotime ordering is unnecessary. Instead of arranging cells according to transcriptomic similarity to one another using a dimensional reduction method (what method did they use??), we used scTOP to compare clones to day 2 undifferentiated cells and day 6 mature cell types. We used the annotations from Weinreb et al. to identify these populations. The resulting reference basis consisted of day 2 undifferentiated progenitors and day 6 neutrophils, monocytes, megakaryocytes, mast cells, eosiniphils, and basophils.

In figure \ref{FIG:5}a, we found the scTOP scores for one particular clonal family that ended up with some sisters that are neutrophils and some that are monocytes. The colors and shapes of the points and colors of the histograms indicate the day the cells were extracted. The undifferentiated scores are indicated on the y-axis of the bottom row, and by the histograms on the lower right. It's clear to see that the cells at day 2 have high undifferentiated scores, and the undifferentiated score decreases as the cells mature -- the histograms for the undifferentiated score move to the left from day 2 to day 6. As the cells mature, their scores for neutrophil and monocytes increase. The graph with neutrophils on the horizontal axis and monocytes on the vertical axis shows that one subset of the population moved towards high-monocyte, low-neutrophil score and the other subset moved towards low-monocyte, high-neutrophil. This indicates that the same progenitor cell can result in either monocytes or neutrophils.

@@

\begin{figure}
	\centering
		\includegraphics[scale=0.75]{figs/scTOP manuscript fig5.pdf}
	\caption{Beyond bench-marking: tracing development in cell-type space with scTOP, using lineage tracing hematopoietic mouse data. (a) {\color{red}} (Describe plots) }
	\label{FIG:4}
\end{figure}

@@

\section*{Acknowledgments}
We acknowledge useful discussions with Jason Rocks, Robert Marsland, and Jason Rocks. The work was funded by a grant to the Boston University Kilachand Multicellular Design Program (to DK and PM) and NIH NIGMS 1R35GM119461 to PM. We also acknowledge programming advice from Alena Yampolskaya.

@@

\subsection{Reference basis construction}
Creating an appropriate reference basis is vital to the accuracy of scTOP. There are many existing scRNA-seq atlases, such as the Mouse Cell Atlas; using them with scTOP requires some amount of curation. For each reference dataset, we took an average over each population corresponding to a given cell type. The average gene expression profiles were preprocessed, and then became the reference profiles for each cell type. Certain cell types from each atlas were dropped for various reasons, such as not including enough cells, being poorly defined (e.g. not enough marker genes to distinguish that cell type from others), etc.

@@

By looking at the scTOP scores for cells whose true type did not match the reference type, we set a score of 0.1 as the lower bound for identifying a type. See figure [REF] for examples of scores where the query type does not match the reference type. In table \ref{table:1}, we labeled cells as "unspecified" if the highest scTOP score was less than 0.1. This provides a useful metric for determining whether a query cell belongs to one of the cell types in the reference basis. If the highest scTOP score is lower than 0.1 for a query cell, this indicates that the reference basis lacks the relevant types or was poorly constructed (see appendix [REF] for a discussion of ill-defined bases).

{\color{red}MY: Using same data as some of the figures in the main paper (ones that don't use the heatmap, since that already shows the irrelevant type scores are low), show that the false positive rate is very low}

@@

\subsection{Robustness of scTOP}
scTOP is consistent between iterations of the algorithm, and it is robust to changes in the reference basis. {\color{red}Discussion and figures showing dropping reference types + see how score distributions change}

@@

The reference cell types should refer to attractor basins. The other part of the assumption is that the reference types should be accurate profiles of those attractor basins. As discussed in appendix [REF: section where \# of cells used is discussed], reference types are created from averaging over cells, and the more cells included in this average the better-sampled the type is. As expected, scTOP gives unreliable results when the reference gene expression levels don't accurately reflect the true cell types' gene expression levels. An important factor in the accuracy of scRNA-seq is the amplification step: after RNA is extracted, it is amplified to improve the sensitivity of sequencing, but not all RNA is amplified at the same rate. This results in inflated counts for some genes but not others. Unique molecular identifiers (UMIs) are vital for reducing the inflating effect of amplification [CITE umi paper]. UMIs tag the RNA before amplification, and are present post-amplification; this makes it possible to see which amplified RNA came from the same original molecule, and prevents uneven double-counting of RNA. We have found that scTOP gives more accurate results when used with reference and query datasets that use UMIs, compared to scTOP results using data that did not use UMIs.

>>>>

\author{Maria Yampolskaya}
\email{mariay@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\author{Michael Herriges}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\author{Laertis Ikonomou}
\affiliation{Department of Oral Biology. University at Buffalo, The State University of New York, Buffalo, NY, USA}
\affiliation{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
\author{Darrell Kotton}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
\author{Pankaj Mehta}
\email{pankajm@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
\affiliation{Biological Design Center, Boston University, Boston, MA 02215, USA}

@@

\section{Introduction}
Humans and other mammals have many specialized cells that arise through a complex developmental process  \cite{zeng_what_2022}. At early embryonic stages, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers \cite{Wolpert_2019}. However, as embryos mature, cellular identity becomes less plastic and more specified through a complex spatiotemporal process involving signaling and environmental cues. The questions of how and why cells end up in one fate or another are fundamental to developmental biology [CITE]. More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineer cells \emph{in vitro}, with important clinical application for human health and disease [CITE].

@@

An ideal analysis algorithm for scRNA-seq data should possess certain desirable properties. 
First, the algorithm should incorporate explicit assumptions that can be kept consistent between analyses. Second, the algorithm should be deterministic, further enhancing reproducibility. Third, the axes in visualization should be interpretable, intuitive, and remain fixed across analysis and datasets. Fourth, the algorithm should measure cell identity in a continuous way instead of assigning discrete identity labels, thus enabling a more subtle and fine-grained analysis of cellular identity and developmental trajectories. Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single cell atlases now being generated [CITE].

@@

\subsubsection{Example 1: Lung lineages}
To demonstrate the efficacy of scTOP in the case where the reference basis and query data come from different sources, we projected mouse lung data from Herriges et al. \cite{herriges_durable_2022} onto the mouse cell atlas. The Herriges et al. dataset  measures gene expression levels of six different specialized lung types from mice used as control samples. See figure \ref{herriges} (a). The reference basis used to find the individual scTOP scores for this data set was derived from the Mouse Cell Atlas. Despite the fact that the Herriges et al. dataset and the Mouse Cell Atlas were created in different laboratories and different sequencing methods, the resulting scTOP scores distinctly separate the cells from the Kotton laboratory. The resulting clusters agree with the annotations of the true cell type from Herriges et al.

\begin{figure}
	\centering
		\includegraphics[scale=0.45]{figs/herriges colored by type.png}
        \includegraphics[scale=0.5]{figs/herriges colored by signature.png}
	\caption{scTOP identifies mouse lung cell types. Using the Mouse Cell Atlas as the reference basis, we show that mouse lung types separate clearly on scTOP score axes. The data points correspond to individual cells, and in the first row of plots, the marker shape and color indicate the "true type" as determined by annotations from Herriges et al. The axes used are scTOP scores for similarity with lung ciliated cells and alveolar types I and II. The second, third, and fourth rows of plots have individual cells colored by the marker genes for the type indicated by the color bars at the end of each row.}
	\label{herriges}
\end{figure}

@@

The mouse brain atlas\cite{yao_taxonomy_2021} sequenced over one million cells and clustered them into 42 subclasses and 101 supertypes. For a reference basis, we reserved a subset of 200 cells from each of the identified subclasses to use as training data. The test data, from which the accuracy score was calculated, was the remaining cells of each type not included in the training data. Ten different cell types are visualized in a grid in figure \ref{mouse_brain} (a). The color of each grid square corresponds to the average scTOP score of cells annotated as the type on the y-axis, with the cell type corresponding to the scTOP dimension on the x-axis. As shown by the distinct blue diagonal, the average scTOP score correctly matches the corresponding annotated cell type.

\begin{figure}
	\centering
    \includegraphics[scale=0.75]{figs/mouse brain.png}
	\caption{scTOP identifies mouse brain cell types. (a) Heatmap showing the average scTOP score for individual cells of the type indicated on the y-axis, compared to the reference types on the x-axis. A subset of cells from the Mouse Brain Atlas are used as the reference basis to query other cells from the same data set. The diagonal indicates that scTOP accurately matches query cells to the true reference type. (b) Histograms showing the distribution of scTOP scores for individual cells of the type labelling the x-axis. The scores shown are of the same type as the sample data. The far left histogram shows the distribtion of Astro scores for Astro cells, and the average of this distribution is indicated by the color of the upper left corner box in heatmap (a).}
	\label{mouse_brain}
\end{figure}

@@

scTOP correctly identifies individual cells, and it can also identify the tissue to which a population of cells belongs. Instead of finding scTOP scores for individual cells, it's possible to group cells and average across them, then find the score for the averaged type. These aggregate pseudo-bulk scores tend to be higher than individual cell scores, since averaging over many cells compensates for the dropout effect. As shown in \ref{tabula_muris}, we found the aggregate scTOP scores for cells from the Mouse Cell Atlas projected onto the reference basis of Tabula Muris organs \cite{schaum_single-cell_2018}. Most cell types were correctly and strongly identified with the annotated tissue of origin. Parenchymal cell types were the most associated with the correct organ, while stromal cells were sometimes misidentified -- for example, the mammary gland endothelial cells. This is consistent with the fact that stromal cells are less specified to each organ. A discussion of the accuracy of scTOP when applied to stromal types versus parenchymal types may be found in appendix \ref{failure_cases}.

\begin{figure}
	\centering
    \includegraphics[scale=0.5]{figs/tabula muris heatmap.pdf}
	\caption{scTOP is able to classify tissues at different resolutions. Aggregate pseudo-bulk samples from the Mouse Cell Atlas are compared to a reference basis consisting of organs from Tabula Muris. The scTOP scores are significantly higher when comparing a cell type with the organ of origin. The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis. The y-axis lists the true types of the queried cells. The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.}
	\label{tabula_muris}
\end{figure}

@@

The CellBench dataset \cite{tian_benchmarking_2019} was designed specifically to benchmark scRNA-seq analysis methods. We used the subset of data pertaining to five cancer lines. 200 cells from each line were used as training data to create the reference basis, then the rest of the cells were input as query data. As shown in figure \ref{cellbench} and table \ref{table:1}, scTOP classifies cell line identity of the test data with 100 percent accuracy.

\begin{figure}
	\centering
		\includegraphics[scale=0.5]{figs/cellbench heatmap.pdf} 
	\caption{scTOP correctly identifies human tissues. The CellBench data set contains human cell cancer lines specifically processed to serve as benchmarking for scRNA-seq algorithms. We take a subset of cells to use as the reference basis in analyzing the rest of the CellBench data. The shade of each bin indicates the average scTOP score for an individual queried cell compared to the cell type indicated on the x-axis. The y-axis lists the true types of the queried cells. The dark diagonal demonstrates the scTOP scores correctly match the queried data to the corresponding true types.}
	\label{cellbench}
\end{figure}

To further prove the power of scTOP in classifying human samples, we analyzed data from the human lung atlas dataset \cite{travaglini_molecular_2020}. This dataset sequenced thousands of human lung cells and identified 58 phenotypic popoulations. For our analysis, we restricted the training and test data to epithelial and stromal types that had at least 200 cells in each type population. The reference basis was created using 80\% of the cells from each population, and the query data consisted of the remaining 20\% of the data. Figure \ref{human lung} shows that the scTOP scores were consistently high in cases where the score type matched the true type of the query.

\begin{figure}
	\centering
		\includegraphics[scale=0.5]{figs/human lung heatmap.pdf}
	\caption{scTOP correctly identifies human tissues. Similar to figure \ref{cellbench}, cells from the Human Lung Atlas are compared to a reference basis constructed from cells from the Human Lung Atlas. The colors of the bins correspond to average scTOP scores of individual cells. Again, the diagonal is the most prominent section, showing that cells of the y-axis types are correctly matched to the reference x-axis identities.}
	\label{human lung}
\end{figure}

@@

\subsubsection{Lung development}
The healthy formation of lungs is vital to mammalian life. For a mammal to breathe and live independently after birth, the final stages of embryonic development must equip it with the ability to exchange gases with its environment. Zepp et al. [CITE] sought to examine the formation of pulmonary alveoli, small sacs where oxygen and carbon dioxide are transferred between the lungs and the air, in the developing mouse. They sequenced mouse lungs between post-conception day 12.5 (E12.5) and postnatal day 42 (P42). As shonw in figure \ref{LungMAP}, we found the scTOP scores for cells annotated by Zepp et al. as type I pneumocytes, type II pneumocytes, and epithelial progenitor cells. The reference basis consisted of adult cell types from the Mouse Cell Atlas and an E12 epithelial progenitor cell type from the Atlas of Lung Development \cite{negretti_single-cell_2021}.

Figure 1I from Zepp et al. conjectures that one begins to see separate alveolar type 1 and type 2 populations between E15.5 and E17.5. According to our plot, at those days there are populations of cells clearly headed towards the two types. There is also a population of cells in between expressing both type 1 and type 2 signatures. Zepp et al. propose these cells are a transitional state similar to the Axin2+ state described by Frank et al. [CITE]. However, as shown in figure \ref{LungMAP genes} (a), the AT1/AT2 cells are not significantly expressing Axin2+. Verheyden and Sun [CITE] summarize the variety of previously-described transitional AT1/AT2 states: alveolar differentiation intermediate (ADI) [CITE Strunz], pre-alveolar type-1 transitional cell state (PATS) [CITE Kobayashi], and damage-associated transient progenitors (DATPs) [CITE Choi]. None of the relevant marker genes for those states were significantly more expressed in the AT1/AT2 cells than either or both of the AT1 or AT2 clusters (figure \ref{LungMAP genes} (b) - (d)). Performing differential gene analysis on the clusters also failed to identify any genes that were uniquely expressed in the AT1/AT2 cluster, indicating the cells in this population are truly in a mixed state.

Because scTOP provides us with reference axes that don't change according to whatever data is included, it is possible to track different data points across time on the same axes. These axes are also easily interpretable: they measure the similarity to a known cell type.

\begin{figure}
	\centering
		\includegraphics[scale=0.7]{figs/LungMAP.png}
	\caption{Visualizing the developing mouse lung with scTOP. (a) scTOP illustrates specification of alveolar type I and type II cells in murine embryonic development. Alveolar cells from Zepp et al. are compared to a reference basis constructed from adult lung cells from Herriges et al. and early epithelial cells from the Atlas of Lung Development. The color and shape of each data point indicates the age of the mouse embryo from which the sample was extracted, from post-conception day 12,5 (E12.5) to 42 days post-birth (P42). (b) The same data as in (a) is displayed as a 2D histogram, with each bin colored corresponding to the average day of the data point(s) falling within that bin. The color trend shows that early cells generally have very low AT1 and AT2 scores, then the AT1 and AT2 scores increase over time.  (c) Scatter plots of the same data as in (a) and (b), now colored by the average expression of AT1 markers genes (left) and AT2 marker genes (right). The scTOP scores of AT1 (AT2) cells are high when the average AT1 (AT2) marker gene expression is high. (d) Scatter plots showing the same data as in the above plots, separated by day to show the combination AT1/AT2 population is most apparent at days P3 and P7.}
	\label{LungMAP}
\end{figure}

@@

To apply LARRY to blood cells, they extracted hematopoietic progenitor cells from mice, barcoded the cells, waited for the progenitors to multiply in primary culture, then performed scRNA-seq at days 2, 4, and 6 post-barcoding. Then, they extracted blood cells at four and six days in cultre. The resulting dataset contains over 300,000 cells, with 2,632 clones spanning time points. Clones belonging to the same clonal family are referred to as "sisters." They then used pseudotime orderings to identify a transcriptomic map to identify lienages of the cells. This map used transcriptomic similarity to place the individual cells accoridng to "pseudotime". With scTOP, creating a pseudotime ordering is unnecessary. Instead of arranging cells according to transcriptomic similarity to one another using a dimensional reduction method (what method did they use??), we used scTOP to compare clones to day 2 undifferentiated cells and day 6 mature cell types. We used the annotations from Weinreb et al. to identify these populations. The resulting reference basis consisted of day 2 undifferentiated progenitors and day 6 neutrophils, monocytes, megakaryocytes, mast cells, eosiniphils, and basophils.

We found scTOP scores for clonal families with sisters that spanned multiple days. The scores for the undifferentiated hematopoietic progenitor stem cell decreased over time, while the scores for the mature types increased. Figure \ref{hem clones}, shows these changes through time for three clonal families.

The clonal family in figure \ref{hem clones} (a) ended up with some sisters that are neutrophils and some that are monocytes. As time passes and the cells mature, their scores for neutrophil and monocytes increase. The graph with neutrophils on the horizontal axis and monocytes on the vertical axis shows that one subset of the population moved towards high-monocyte, low-neutrophil score and the other subset moved towards low-monocyte, high-neutrophil. This indicates that the same progenitor cell can result in either monocytes or neutrophils.

scTOP can reveal differentiation pathways without the need for pseudotime inference. Pseudotime inference and other dimensional reduction methods place cells next to one another, which can bias results. It's difficult to draw conclusions about certain genes being expressed at different points in pseudotime; patterns may arise that suggest a continuous progression of differentiation, but the progression may be the result of an algorithm inferring that cells which express similar genes should be next to one another in pseudotime, creating a progression by assumption. scTOP avoids this issue by comparing cells to a reference instead of grouping them by similarity to one another. Weinreb et al. claim that the dimensional reduction shown in figure 4 (a) indicates there are two pathways of monocyte differentiation, one that expresses dendritic cell markers and one that expresses neutrophil markers. We used scTOP to visualize all the in vitro clones in terms of similarity to neutrophils, monocytes, and progenitors (\ref{hem genes}). This allows us to see which cells are more likely to be neutrophils that exhibit some monocyte-like genes rather than neutrophil-like monocytes. However, one limitation of scTOP is that it requires sufficient cells to create a reliable reference from; since there were too few dendritic cells to serve as reference, it was not possible to calculate dendritic cell scores.

@@

\begin{figure}
	\centering
		\includegraphics[scale=0.75]{figs/hem. clone families.png}
	\caption{Tracing hematopoietic cells differentiation with scTOP. (a), (b), (c) display scatter plots and distributions of scTOP scores for individual in vitro clone families from Weinreb et al. The scatter plots show the x-axis scTOP score compared to the y-axis scTOP score for individual cells, while the distributions show the scTOP scores of the type indicated on the x-axis of the corresponding column. The color and shape of each marker (and color of each distribution) indicate the days in culture at which the cells were extracted. (a) A clone family where some clones end up becoming neutrophils and some become monocytes. At day 2, the 3 sister cells have low neutrophil and monocyte scores, and high progenitor scores. At days 4 and 6, the progenitor scores decrease and a portion of the cells increase in monocyte scores while others increase in neutrophil scores. (b) A clone family where all of the cells end up becoming neutrophils. (c) A clone family where the descendant cells become neutrophils, eosiniphils, and basophils.}
	\label{hem clones}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[scale=0.75]{figs/hem progenitor hists.png}
	\caption{Tracing differentiation of hematopoietic cells from progenitor to mature type. The plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al., with the x-axis of each one indicating the scTOP score for hematopoietic progenitor type and the y-axis showing the score for a mature type. Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin. Early cells have high progenitor scores and low mature type scores. As time passes cells end up with high scores in mature types.}
	\label{hem progenitor}
\end{figure}

\begin{figure}
	\centering
		\includegraphics[scale=0.59]{figs/hem mature hists.png}
	\caption{Comparing differentiation of hematopoietic cells between mature types. As in figure \ref{hem progenitor}, the plots shown here are 2D histograms of all of the in vitro clone families from Weinreb et al. The x-axis and y-axis of each histogram indicate the scTOP score of the labelled mature type. Each bin of the histogram is colored by the average day in culture of the data points that fall within the bin.  Some pairs of types have cells with high scores of both types. This is apparent in the neutrophil-monocyte histogram, which presents a continuous spectrum of scores from the horizontal axis to the vertical. Other pairs of types only have cells expressing one type or the other, as in the L-shaped neutrophil-megakaryocyte histogram.}
	\label{hem mature}
\end{figure}

@@

\section*{Acknowledgments}
We acknowledge useful discussions with Jason Rocks and Robert Marsland. The work was funded by a grant to the Boston University Kilachand Multicellular Design Program (to DK and PM) and NIH NIGMS 1R35GM119461 to PM. We also acknowledge programming support from Alena Yampolskaya.

@@

\subsection{Reference basis construction}\label{basis construction}
Creating an appropriate reference basis is vital to the accuracy of scTOP. There are many existing scRNA-seq atlases, such as the Mouse Cell Atlas; using them with scTOP requires some amount of curation. For each reference dataset, we took an average over each population corresponding to a given cell type. The average gene expression profiles were preprocessed, and then became the reference profiles for each cell type. Certain cell types from each atlas were dropped for various reasons, such as not including enough cells, being poorly defined (e.g. not enough marker genes to distinguish that cell type from others), etc.

@@

By looking at the scTOP scores for cells whose true type did not match the reference type, we set a score of 0.1 as the lower bound for identifying a type. As shown in figure \ref{robustness hists} (c) and (f), when the reference type does not match the sample type the scores for individual cells fall well within [-0.1, 0.1], even when most of the reference basis is missing.

In table \ref{table:1}, we labeled cells as "unspecified" if the highest scTOP score was less than 0.1. This provides a useful metric for determining whether a query cell belongs to one of the cell types in the reference basis. If the highest scTOP score is lower than 0.1 for a query cell, this indicates that the reference basis lacks the relevant types or was poorly constructed (see appendix \ref{failure_cases} for a discussion of ill-defined bases).

@@

\subsection{Robustness of scTOP}
scTOP is consistent between iterations of the algorithm, and it is robust to changes in the reference basis. As shown in figure \ref{robustness hists}, many scTOP scores do not change significantly even when most of the reference basis is removed. Generally, scores for cells that match the reference type do not change much even when only 25\% of the original basis is retained (figure \ref{robustness hists} (d), (e)). The largest difference occurs for cells that are similar to the reference type, such as the AT1 score for AT2 cells (figure \ref{robustness hists} (g)). These scores tend to increase, because removing other types from the basis reduces the ability to de-correlate effectively. In other words, scTOP is more likely to confuse similar types when it has fewer reference types to compare. This effect is most apparent when relevant types are removed from the basis. Figures \ref{robustness hists} (j) and (k) show that all lung types score higher for AT1 and AT2 when only lung alveolar types are included in the basis. Figures \ref{robustness hists} (c) and (f) show that irrelevant scores, such as bladder and thymus cells, do not change much even when large portions of the basis are removed. scTOP does not confuse sample cells for types that are dissimilar even with a small reference basis because there is no need to de-correlate between types that are naturally not correlated.

@@

The reference cell types should refer to attractor basins. The other part of the assumption is that the reference types should be accurate profiles of those attractor basins. As discussed in appendix \ref{basis construction}, reference types are created from averaging over cells, and the more cells included in this average the better-sampled the type is. As expected, scTOP gives unreliable results when the reference gene expression levels don't accurately reflect the true cell types' gene expression levels. An important factor in the accuracy of scRNA-seq is the amplification step: after RNA is extracted, it is amplified to improve the sensitivity of sequencing, but not all RNA is amplified at the same rate. This results in inflated counts for some genes but not others. Unique molecular identifiers (UMIs) are vital for reducing the inflating effect of amplification [CITE umi paper]. UMIs tag the RNA before amplification, and are present post-amplification; this makes it possible to see which amplified RNA came from the same original molecule, and prevents uneven double-counting of RNA. We have found that scTOP gives more accurate results when used with reference and query datasets that use UMIs, compared to scTOP results using data that did not use UMIs.
